---
title: "QC: Raw Data"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
bibliography: references.bib
link-citations: true
---

```{r setup, echo = FALSE}
conda_pre <- system2("echo", "$CONDA_PREFIX", stdout = TRUE)
if (conda_pre != "") {
  conda_lib_path <- file.path(conda_pre, "lib", "R", "library")
  if (!dir.exists(conda_lib_path)) conda_lib_path <- NULL
  prev_paths <- .libPaths()
  paths_to_set <- unique(c(conda_lib_path, prev_paths))
  .libPaths(paths_to_set)
}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.height = 8, fig.width = 10, fig.align = 'center'
)
```

```{r packages}
library(tidyverse)
library(glue)
library(yaml)
library(here)
library(reactable)
library(pander)
library(ngsReports)
library(scales)
library(htmltools)
library(Polychrome)
myTheme <- theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 13)
  )
```

```{r load-data}
config <- read_yaml(here::here("config/config.yml"))
samples <- read_tsv(here::here(config$samples)) %>% 
  bind_rows(
    tibble(
      accession = unique(.$input), target = "Input"
    )
  )
n <- length(samples$accession)
pal <- createPalette(n, c("#2A95E8", "#E5629C"), range = c(10, 60), M = 100000)
names(pal) <- samples$accession
colours <- scale_colour_manual(values = pal)
qc_path <- here::here(config$paths$qc, "raw")
rel_qc_path <- ifelse(
  grepl("docs", qc_path), gsub(".+docs", ".", qc_path),
  gsub(here::here(), "..", qc_path)
)
fl <- file.path(qc_path, glue("{samples$accession}_fastqc.zip")) %>% 
  setNames(samples$accession)
rawFqc <- FastqcDataList(fl[file.exists(fl)])
names(rawFqc) <- names(fl)[file.exists(fl)]
```

## Introduction

This page provides summary statistics and diagnostics for the raw data as obtained at the commencement of the workflow.
Most data was obtained from [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/) and parsed using the Bioconductor package ngsReports [@Ward2020-fn].
FastQC reports for `r length(rawFqc)` files were found.

A conventional MultiQC report [@Ewels2016-ix] can also be found [here](`r rel_qc_path`/multiqc.html)

## Data Summary {.tabset}

```{r tbl-read-totals}
div(
  class = "table",
  div(
    class = "table-header",
    htmltools::tags$caption(
      htmltools::em(
        "Library sizes with links to all raw FastQC reports"
      )
    )
  ),
  readTotals(rawFqc) %>% 
    mutate(Filename = str_remove_all(Filename, ".(fast|f)q.gz")) %>% 
    left_join(samples, by = c("Filename" = "accession")) %>% 
    dplyr::select(-input) %>% 
    setNames(str_replace_all(names(.), "_", " ")) %>% 
    setNames(str_to_title(names(.))) %>% 
    reactable(
      sortable = TRUE, resizable = TRUE,
      showPageSizeOptions = TRUE,
      columns = list(
        Filename = colDef(
          cell = function(value) htmltools::tags$a(
            href = file.path(rel_qc_path, glue("{value}_fastqc.html")), 
            target = "_blank", 
            value
          ),
          html = TRUE
        )
      ),
      defaultColDef = colDef(format = colFormat(separators = TRUE))
    )
)
```


### FastQC Status

```{r plot-summary, fig.cap = "*Summary of Pass/Warn/Fail status from all samples*"}
plotSummary(rawFqc) + myTheme
```


### Read Totals


```{r plot-read-totals, fig.cap = "*Raw library sizes showing estimated duplication levels*"}
p <- plotReadTotals(rawFqc, divBy = 1e6) + myTheme
p$data <- left_join(p$data, samples, by = c("Filename" = "accession"))
p + facet_grid(target + treatment ~ ., scales = "free", space = "free")
```


## QC Diagnostics

### Quality Scores

```{r plot-base-quals, fig.cap = "*Mean Quality Scores are each position along the reads*"}
plotBaseQuals(rawFqc, heat_w = 20, dendrogram = TRUE, usePlotly = TRUE, text = element_text(size = 13))
```

### Sequence Content Heatmap

```{r plot-seq-content, fig.cap = "*Sequence content along each read*"}
plotSeqContent(rawFqc, usePlotly = TRUE, dendrogram = TRUE, heat_w = 20, text = element_text(size = 13))
```

### Sequence Content Residuals {.tabset}

#### All Libraries

```{r plot-seq-residuals, fig.cap = "*Residuals obtained when subtracting mean values at each position*"}
plotSeqContent(rawFqc, plotType = "residuals", scaleColour = colours) + myTheme
```

```{r plot-tgt-seq-residuals}
d <- here::here("docs", "assets", "raw_fqc")
if (!dir.exists(d)) dir.create(d, recursive = TRUE)
h <- knitr::opts_chunk$get("fig.height")
w <- knitr::opts_chunk$get("fig.width")
htmltools::tagList(
  samples %>% 
    dplyr::filter(accession %in% names(rawFqc)) %>% 
    split(.$target) %>% 
    setNames(NULL) %>% 
    lapply(
      function(x) {
        tgt <- unique(x$target)
        fqc <- rawFqc[x$accession]
        p <- plotSeqContent(fqc, plotType = "residuals", scaleColour = colours) +
          ggtitle(glue("Sequence Content Residuals: {tgt} libraries only")) +
          myTheme
        png_out <- file.path(
          d, glue("seq-content-residuals-{tgt}.png")
        )
        href <- str_extract(png_out, "assets.+")
        png(filename = png_out, width = w, height = h, units = "in", res = 300)
        print(p)
        dev.off()
        cp <- htmltools::em(
          glue(
            "
            Residuals obtained when subtracting mean values at each position for 
            {tgt} libraries only.
            "
          )
        )
        htmltools::div(
          htmltools::div(
            id = glue("plot-seq-residuals-{tgt}"),
            class = "section level4",
            htmltools::h4(tgt),
            htmltools::div(
              class = "figure", style = "text-align: center",
              htmltools::img(src = href, width = "100%"),
              htmltools::tags$caption(cp)
            )
          )
        )
      }
    )
)
```

### GC Content {.tabset}

#### All Libraries

```{r plot-gc-content, fig.cap = "*GC content distributions for all libraries*"}
plotGcContent(
  rawFqc, usePlotly = TRUE, plotType = "line", theoreticalGC = FALSE, 
  scaleColour =  colours, plotlyLegend = TRUE, text = element_text(size = 13)
)
```

```{r plot-tgt-gc-content}
h <- knitr::opts_chunk$get("fig.height")
w <- knitr::opts_chunk$get("fig.width")
htmltools::tagList(
  samples %>% 
    dplyr::filter(accession %in% names(rawFqc)) %>% 
    split(.$target) %>% 
    setNames(NULL) %>% 
    lapply(
      function(x) {
        tgt <- unique(x$target)
        fqc <- rawFqc[x$accession]
        p <- plotGcContent(
          fqc, plotType = "line", theoreticalGC = FALSE, scaleColour = colours
        ) +
          ggtitle(glue("GC Content Distribuions: {tgt} libraries only")) +
          myTheme
        png_out <- file.path(
          d, glue("gc-content-{tgt}.png")
        )
        href <- str_extract(png_out, "assets.+")
        png(filename = png_out, width = w, height = h, units = "in", res = 300)
        print(p)
        dev.off()
        cp <- htmltools::em(
          glue("GC content distributions for {tgt} libraries only.")
        )
        htmltools::div(
          htmltools::div(
            id = glue("plot-gc_content-{tgt}"),
            class = "section level4",
            htmltools::h4(tgt),
            htmltools::div(
              class = "figure", style = "text-align: center",
              htmltools::img(src = href, width = "100%"),
              htmltools::tags$caption(cp)
            )
          )
        )
      }
    )
)
```

### Sequence Lengths

```{r plot-seqlengths, fig.cap = "*Cumulative Sequence Length distributions*"}
plotSeqLengthDistn(
  rawFqc, usePlotly = TRUE, plotType = "cdf", scaleColour = colours, 
  plotlyLegend = TRUE, text = element_text(size = 13)
)
```


### Duplication Levels

```{r plot-dup-levels, fig.cap = "*Duplication levels as estimated by FastQC*"}
plotDupLevels(
  rawFqc, usePlotly = TRUE, dendrogram = TRUE, heat_w = 20, 
  text = element_text(size = 13)
)
```

### Adapter Content 

```{r plot-adapter-content, fig.cap = "*Total Adapter Content across all samples. If the total is below 0.1% an empty plot will be shown.*"}
plotAdapterContent(
  rawFqc, usePlotly = TRUE, plotType = "line", scaleColour = colours, 
  text = element_text(size = 13), plotlyLegend = TRUE
)
```


## References

<div id="refs"></div>

<br>
<button type="button" class="btn btn-default btn-sessioninfo" data-toggle="collapse" data-target="#sessioninfo" style="display: block;">
<span class="glyphicon glyphicon-info-sign" aria-hidden="true"></span> Session information
</button>
</p>
<div id="sessioninfo" class="collapse">
```{r session-info, echo=FALSE}
pander::pander(sessionInfo())
```
</div>
